---
title: "Title of your project"
author: "Your Name"
date: today
abstract: fuu 
format: html
---

This is a Quarto document. 
You can combine text and code here.
Write **bold** and *emphasized* text like this.
Include [links](https://petrpajdla.github.io/stat4arch) like this.
To create a new paragraph, leave a line empty.

Include code blocks like this:

```{r}
#| label: code-block-title

# with single hash symbol, include comments 
# in the code blocks as you are used to from scripts
1 + 1
```

Next is a template for your own project report...


## Abstract

Short abstract describing what is the data set about and what you are focusing on in the analysis (one paragraph of text at most).

**Keywords**

Include several useful keywords.


## Introduction

Mention what is the goal of your analysis, why did you choose this data set etc. in a short introduction.
Do not forget to include some research questions.

Load your libraries here.

```{r}
#| label: libs

# some examples are here:
# library(dplyr)
# library(here)
# library(ggplot2)
```

## Data

Briefly introduce your dataset. 
Where does it come from? What does it contain? 
What are the variables and observations? 
How did you get the dataset? 
Where is it originally published (include citations, links etc.) 

Read your data in the next code block:

```{r}
#| label: data

# example:
# my_dataset <- read.csv(here("path/to/your/dataset.csv"))
```

Explore the data set and try to answer following questions using code:

- How many variables are there?
- How many rows does the data set have?
- What are the names of the variables?
- What data types are there in the data set?

```{r}
#| label: data-exploration

# ncol(my_dataset)
# nrow(my_dataset)
# and so on, you know the basic functions...
```


## Analysis

Focus on the analysis of your data set. 
Are there any quantitative and qualitative variables? 
Show answers to questions like these on graphs and using various numerical summaries or tests:

- How are the variables distributed?
- What are their central points and how are they spread? What does it mean?
- Is any of the quantitative variables distributed normally?
- What are the relationships between the variables?
- If your points, sites etc. have coordinates, show them on a simple map.

You know how to create code blocks now, so go for it!
**Warning** if you copy the code blocks, always change their labels, or omit them completely, because code block labels must be unique throughout the document!
For more help and options, have a look here: <https://quarto.org>.


## Summary

Briefly sum up and interpret what you managed to achieve in your project. 
Did you learn anything new? Did anything surprise you?

