---
title: "Data transformation"
format:
  html: default
  revealjs:
    fontsize: 24px
    scrollable: true
    embed-resources: true 
    output-file: slides_transformation.html
    footer: "[AES_707](https://petrpajdla.github.io/stat4arch/) *Statistics seminar for archaeologists* | [Data transformations](./transformation.html)"
execute:  
  echo: true
  warning: false
  cache: true
---

```{r}
#| include: false
library(dplyr)
library(here)
library(ggplot2)
df_dartpoints <- read.csv(here("lect/w02/data/dartpoints.csv"))
df_sample <- df_dartpoints |> 
  select(Name, Length, Width, Weight, Base.Sh)
```


## Prerequisites

### Packages:
- `here`
- `dplyr`

### Data:
- dartpoints.csv

### Start:
 - open your R Project, create a new script file
 - load library `here`, install and load library `dplyr`
 - import data "dartpoints.csv

::: fragment
### Hints:
- `install.packages()`, `library()`, `here()`, `read.csv()`
:::

## Course overview

In this course, you will learn how to:

- select specific columns or remove them -- `select()`
- rename your columns or change their position -- `rename()`, `relocate()`
- create new columns and change the values based on different conditions -- `mutate()`, `case_when()`
- filter specific rows based on various conditions -- `filter()`
- order your rows from lowest to highest values (or vice versa) -- `arrange()`
- group your data and calculate different summary statistics -- `summarise()`, `group_by()`
- work with different functions more effectively with the pipe operator -- `|>`

::: fragment
At the end, you will be able to create and understand such a crazy code:

```{r}
#| eval: FALSE
df_sample |> 
  group_by(Name) |> 
  summarise(
    mean_length = round(mean(Length),1),
    median_length = round(median(Length),1),
    sd_length = round(sd(Length),1),
    n_name = n()
  ) |> 
  mutate(
  percent_name = round(n_name/sum(n_name)*100, 1)
  ) |> 
  ggplot()+
  aes(x = Name, y = sd_length)+
  geom_col(fill = "steelblue")+
  theme_light()
```
:::
## package `dplyr`
- we will work with the package `dplyr` which uses tidy data logic (one variable = one column, one observation = one row,...)
- a big advantage of this package is that it has more intuitive and human readable syntax

::: fragment
- let's say we want to analyse only those dartpoints that have name "Travis"
- in R base, the code to select only dartpoints of that type, would look like this:
:::


::: fragment
```{r}
#| eval: false
df_dartpoints[df_dartpoints$Name == "Travis",]
```

:::

::: fragment

This is a solution with the function `filter()` from the `dplyr` package:

```{r}
#| eval: false
filter(df_dartpoints, Name == "Travis")
```
:::
::: fragment
Similarly, you can also select different columns by their name:

R base solution:

```{r}
#| eval: false
df_dartpoints[, c("Name", "Length", "Width", "Weight")]
```
:::
::: fragment
- and this is the `dplyr` solution with function `select()`:

```{r}
#| eval: false
select(df_dartpoints, Name, Length, Width, Weight)
```
:::


## pipe operator - `|>`
- `|>` allows you to combine various functions into one code chunks, which makes your code shorter and easier to read

- syntax:

`dataframe |> function_1(variable, condition) |> function_2(variable, condition)`

::: fragmnent
So you can combine functions `select()`, `filter()` and `head()` :

```{r}
df_dartpoints |> 
  select(Name, Length, Width, Weight) |> 
  filter(Name == "Travis") |> 
  head(4)
```

- notice you don´t need to repeat the name of the dataframe into every function arguments, since you already specified it in the begining of the "pipe"
:::


## `select()` - more details

```{r}
df_dartpoints |> 
  select(Name, Length, Width, Weight, Blade.Sh) |> 
  head(4)
  
```
::: fragment
```{r}
df_dartpoints |> 
  select(Name, Length:Weight) |> 
  head(4)
```
:::
::: fragment
 
You can exclude specific columns by `!`

```{r}
df_dartpoints |> 
  select(!Name) |> 
  head(4)
```
::: 
::: fragment

```{r}
df_dartpoints |> 
  select(!Name:Quad) |> 
  head(4)
```
:::
::: fragment
- selecting columns with numeric variables:

```{r}
df_dartpoints |> 
  select(where(is.numeric)) |> 
  head(4)
```
:::

## `rename()`
- you can change the names of your columns by `rename()` function
- syntax:

  `rename(new_name = old_name)`

::: fragment
```{r}
df_dartpoints |> 
  rename(
    typ_sipky = Name,
    delka_sipky = Length,
    sirka_sipky = Width
  ) |> 
  head(4)
```


:::

## `relocate()`
- `relocate()` will defautly move your chosen column to the front:
```{r}
df_dartpoints |> 
  rename(
    typ_sipky = Name,
    delka_sipky = Length,
    sirka_sipky = Width
  ) |> 
  relocate(delka_sipky, sirka_sipky) |> 
  head(4)
```
- note that you must already be working with the new variable names.

::: fragment
- you can manualy specify the position of relocated columns by aruments `.before = ` or `.after = `
```{r}
df_dartpoints |> 
  rename(
    typ_sipky = Name,
    delka_sipky = Length,
    sirka_sipky = Width
  ) |> 
  relocate(
    delka_sipky, sirka_sipky, .after =  typ_sipky
    ) |> 
  head(4)
```
- note the dot before arguments `.after` and `.before` - it indicates to R that those are not objects or variables, but arguments of a function

:::

## Before we continue:
- you can also create a new object from your selection with `<-`:

```{r}
df_sample <- df_dartpoints |> 
  select(Name, Length, Width, Weight, Base.Sh)
head(df_sample, 4)
```

From now on, we will continue with this dataframe, in order to reduce the number of the columns.

## `mutate()`
- this function creates new columns that are calculated from existing ones (i.e. the new columns are functions of existing variables).

- syntax:
`dataframe |> mutate(variable_name = function())`

::: fragment
- here is an example of creating a new column whose values are the multiplication of length by width
```{r}
df_sample |> 
  mutate(
    square_size = Length*Width
    ) |> 
  head(4)
```
:::
::: fragment

- note what argument `.keep = "used"` does:
```{r}
df_sample |> 
  mutate(
    square_size = Length*Width,
    .keep = "used"
    ) |> 
  head(4)
```
- we will explore the function `mutate()` in more detail later

:::

## `filter()`

- for specifying conditions for filtering, `dplyr` uses the following logical and mathematical operators: ==, !=, \<, \>, \>=, \<=, &, \|, \%in%,  etc (use `?dplyr::filter` for more details)

::: fragment
- here, we will use the symbol `>` to filter out dart points with a length greater than 80 mm

```{r}
df_sample |> 
  filter(Length > 80)
```
:::

::: fragment
- and here we use `==` to choose only those dartpoints which are of type "Travis"

```{r}
df_sample |> 
  filter(Name == "Travis")
```
:::

::: fragment
- alternatively, you can exclude all points of a type "Travis" by negation `!=`

```{r}
df_sample |> 
  filter(Name != "Travis") |> 
  head(4)
```
:::

::: fragment
- add `&` if you want to filter with more than one condition, for example here we will filter all points which are type "Wells" **AND** are heavier than 10 grams

```{r}
df_sample |> 
  filter(Name == "Wells" & Weight > 10)
```
:::

::: fragment
- {{< bi wrench-adjustable >}} Task: instead of `&` try the operator `|` (**OR**) and see how the result differs
:::

## Filtering based on a vector
- you can make your code less complicated when you create vector from desired values and then filter all observations which fall into that vector by using operator `%in%`

```{r}
two_types <- c("Pedernales", "Ensor")

df_sample |> 
  filter(Name %in% two_types) |> 
  head(4)
```

## `arrange()`

- this function will order your observations from the lowest to highest (or vice versa)

```{r}
df_sample |> 
  arrange(Length) |> 
  head(4)
```


```{r}
df_sample |> 
  arrange(desc(Length)) |> 
  head(4)
```

## `distinct()` and `count()`
- try to find out what these functions are doing:

```{r}
df_sample |> 
  distinct(Name, Base.Sh) |> 
  head(10)
```

```{r}
df_sample |> 
  count(Name, Base.Sh, sort = TRUE) |> 
  head(4)
```

## `summarise()`
- this function is used for calculating basic summary stastics from the existing variables
- syntax:
 ` summarise(new_variable = function(variable))`
 - in this case, the mean and sd values are calculated for whole dataframe:
```{r}
df_sample |> 
  summarise(
    mean_lenght = mean(Length),
    sd_length = sd(Length)
  )
```

## `group_by()`
- if you want to calculate summary statistic for each dartpoint type, you have to use function `group_by` which will create groups of rows based on chosen variable, in this case "Name"
```{r}
df_sample |> 
  group_by(Name) |> 
  summarise(
    mean_lenght = mean(Length),
    sd_length = sd(Length),
    n = n()
  ) |> 
  arrange(desc(n))
```

::: fragment
```{r}
df_sample |> 
  group_by(Name, Base.Sh) |> 
  summarise(
    mean_lenght = round(mean(Length),2),
    sd_length = round(sd(Length),2),
    n = n()
  ) |> 
  arrange(Name) |> 
  head(7)
```
:::

## back to `mutate()`

### creating new values based on different conditions


```{r}
df_sample |> 
  mutate(
    size_cathegory = case_when(
      Length > 70 ~ "big_stone",
      Length < 30 ~ "small_stone",
      .default = "ok_stone"
    )
  ) |> 
  arrange(desc(Length)) |> 
  head(10)
```

## back to `mutate()`
### changing existing values based on combination of two conditions
- imagine we want to create a new type of dartpoints from the Pedernales that are longer than 80 mm
- we don't need to create a new column, but only change the values in the existing one:

```{r}
df_sample |> 
  mutate(
    Name = case_when(
      Name == "Pedernales" & Length > 80 ~ "Big_Pedernales",
      .default = Name
    )
  ) |> 
  arrange(desc(Length)) |> 
  head(4)
```

- note that if we want to leave the other values of the variable "Name" unchanged, we do so by specifying `.default = Name`

## Combining plenty of functions into one huge code chunk: 

```{r}
df_sample |> 
  group_by(Name) |> 
  summarise(
    mean_length = round(mean(Length),1),
    median_length = round(median(Length),1),
    sd_length = round(sd(Length),1),
    n_name = n()
  ) |> 
  mutate(
  percent_name = round(n_name/sum(n_name)*100, 1)
  ) |> 
  arrange(desc(n_name))
  
```

::: fragment
- we can even include `ggplot()` into the pipeline:

```{r}
df_sample |> 
  group_by(Name) |> 
  summarise(
    mean_length = round(mean(Length),1),
    median_length = round(median(Length),1),
    sd_length = round(sd(Length),1),
    n_name = n()
  ) |> 
  mutate(
  percent_name = round(n_name/sum(n_name)*100, 1)
  ) |> 
  ggplot()+
  aes(x = Name, y = sd_length)+
  geom_col(fill = "steelblue")+
  theme_light()
```

## Saving your result

```{r}
df_result <- df_sample |> 
  group_by(Name) |> 
  summarise(
    mean_length = round(mean(Length),1),
    median_length = round(median(Length),1),
    sd_length = round(sd(Length),1),
    n_name = n()
  ) |> 
  mutate(
  percent_name = round(n_name/sum(n_name)*100, 1)
  ) |> 
  arrange(desc(n_name))
```


```{r}
#| eval: false
write.csv(df_result, here("results/dartpoints_result.csv"), row.names = F)
```

